
---
title: "COVID-19 Data Analysis and Insights"
author: "Chris Bunker"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
---

# Introduction

The Covid-19 Pandemic was one of the most significant events of the 21st 
century. It affected people from across the globe, regardless of socio-economic
status. The pandemic was believed to have originated in Wuhan, China and after
spreading rapidly, was officially designated a pandemic by the World Health
Administration.

As of April 13, 2024, over 7 million people have died as a direct result of 
Covid-19, with the disease affecting the elderly and immune-compromised the
most significantly.

It has long been known that diseases do not affect different populations 
equitably. Countries considered to be "second-world" or "third-world" 
historically have faced extreme disparities in medical treatment when compared
to those considered "first-world". For instance, in the United States there was
an estimated 9,615 cases of TB in 2023, which is a significant number. However, 
in Indonesia, a country with a population around 50 million people less than the
US, experienced 821,000 thousand cases of TB in 2023.

The purpose of my data analysis using a publicly available Covid dataset
from https://ourworldindata.org/coronavirus is to explore the relationship 
between economic status and country, to see if the trend that has historically
existed, exists with Covid-19 as well.

However, I will also be keeping in mind important information, such as the 
bias of data collection, where data collection methods are more prominent and
tend to be more accurate in first world nations. Also keeping in mind the bias
to for research and medical treatment to support first world nations rather
than third world nations.

Below are references used in this passage: 

Northwestern Medicine. (n.d.). COVID-19 pandemic timeline. Retrieved from 
https://www.nm.org/healthbeat/medical-advances/new-therapies-and-drug-trials/
covid-19-pandemic-timeline#:~:text=By%20March%202020%2C%20the%20World,
COVID%2D19%20outbreak%20a%20pandemic.

Worldometer. (n.d.). Coronavirus death toll.
Retrieved from https://www.worldometers.info/coronavirus/coronavirus-death-toll/

Centers for Disease Control and Prevention (CDC). (2023).
Mortality in the United States, 2021.
Retrieved from https://www.cdc.gov/mmwr/volumes/73/wr/mm7312a4.html

USAID. (2024). Indonesia tuberculosis roadmap narrative. 
Retrieved from https://www.usaid.gov/sites/default/files/2024-01/
Indonesia_TB_roadmap_narrative-22_edits_508.pdf
---

# Data and Methods

The data contained Covid related information on around 250 countries throughout
the world. Depending on the specific analysis, some of these countries did not
have sufficient data while most did. 

My analysis focused most heavily on total cases per million, total deaths per
million, GDP Per Capita, and the Country where it occurred. For each of these
countries I assigned a GDP Status category based upon GDP Per Capita where the
lowest group had less than $10,000 USD GDP Per Capita, the second lowest having 
greater than $10,000, the third greater than $20,000, and the highest group
having more than $35,000 USD GDP Per Capita.

## Data Sources

The analysis is based on the following data:
- **COVID-19 Data**: Sourced from [Our World in Data](https://ourworldindata.org/coronavirus).
- **Geospatial Data**: Administrative boundaries from the Natural Earth dataset.


## Methods

The visualizations in this report were generated using R and include:
- Choropleth maps for geographic data visualization.
- Scatterplots and histograms to explore relationships and distributions.
- Bar plots to highlight economic indicators.

---

# Analysis and Results

## 1. Economic Indicators: GDP Distribution by Country

The plot below simply shows the GDP Per Capita for various countries.

```{r}
readRDS("history_plot.rds")
```

*Key Observations*: [As we can see, the majority of all countries have a GDP Per
Capita under 10K USD, amounting to almost 50% for all of the data. The other 
three categories make up the rest.]

---

## 2. Mortality Rates: Distribution of Reported COVID-19 Deaths

This histogram shows the distribution of deaths caused by Covid-19 by
April 25th, 2022 per million people. 

```{r}
readRDS("histogram_plot.rds")
```

*Key Observations*: [As we can see, most countries appear to have deaths under
2,000 per million people. However, there is a significant right skew to the data
with one value even greater than 6,000. Further analysis will reveal if economic
status has a correlation to these outliers. Furthermore, I am skeptical of the 
data centered near and at zero and wonder instead if this is a problem with
reporting.]

---

## 3. Demographics: Relationship Between Median Age and COVID-19 Cases

This scatter plot looks at age, total cases per million people, and economic
status for individual countries. Each dot represents and individual country. The
median age represents the median age of the country's population.

```{r}
readRDS("final_plot.rds")
```

*Key Observations*: [Surprisingly, we see that high GDP countries have higher 
rates of Covid-19. This may be due to factors unique to wealthy individuals, 
such as the tendency to travel as well as the nature of Covid-19 which like 
other seasonal viruses such as the flu, tends to increase in transmission during
the winter when people are stuck inside. Countries closer to the equator do not
experience this issue, so as well as tending to have a lower GDP, also have a 
layer of protection from Covid-19. Like many viruses, it is also possible that 
Covid-19 simply performs better in one climate than the other. In this case, the 
protective factors for impoverished countries close to the equator tend to 
overcome the negating factors which historically have resulted in high levels
of TB and other diseases.]

---

## 4. Geographic Impact: COVID-19 Deaths in the Americas

This choropleth map highlights the geographic distribution of COVID-19 deaths 
across the Americas, focusing on Central and South America.

```{r}
readRDS("map_americas.rds")
```

*Key Observations*: [The chloropeth map reveals a somewhat interesting finding:
economic status does not appear to play a significant role in overall deaths 
caused by Covid-19. The global south encapsulates many third world countries, 
but as we can see, the death rates are comparable to that of the US and Canada
and we do not see significant differences between wealthier and poorer countries
in South America.

However, what we do see is one clear outlier: Peru. After further research, I 
found that Peru had the world's highest death rate. At the time Covid hit Peru 
in early 2020, Peru was already being hit by a massive wave of Dengue fever,
rendering the medical sector incapable of adequately responding to Covid. 
Furthermore, Peru's medical system ran out of oxygen which was necessary to 
keep patients in a severe state alive. Other factors such as the informal economy
and housing crisis in Peru expedited the spread of the virus as well.

Northwestern Medicine. (n.d.). COVID-19 pandemic timeline. Retrieved from https://www.nm.org/healthbeat/medical-advances/new-therapies-and-drug-trials/
covid-19-pandemic-timeline#:~:text=By%20March%202020%2C%20the%20World,COVID%2D]

---

# Discussion

My analysis and visualization reveals that Economic Status can play a role in
cases and deaths from Covid-19. However, the relationship is not linear, and
more descriptive than causal.

For instance, a lack of medical supplies and overcrowding definitely were a 
significant cause of deaths in Peru. However, this is not a trend that is 
uniform throughout the data.

Rather, the seasonality of the virus and increased transmission during the 
winter may likely be a significant factor working against so called "first 
world nations". As we saw in the scatter plot, first world nations tended to
represent a significant portion of the global Covid-19 cases. Impoverished 
countries, with a skew towards the global south, appear to have a layer of 
protection due to their lack of seasons. However, other factors such as 
overcrowding and lack of medical supplies and treatment tend to balance out
with northern countries. This results in a chloropeth (map) that shows a 
somewhat indiscriminate trend, where economic status and deaths caused by Covid
do not seem correlated.

---

# Conclusion

Overall, there is not significant evidence to suggest that economic status is 
a driving force for the number of cases or deaths caused by Covid-19, except in
the case of Peru where this is significant evidence suggesting that a lack of
medical supplies were directly responsible for the largest death toll by 
Covid-19 globally.

---

# References

1. [Our World in Data: COVID-19](https://ourworldindata.org/coronavirus)
2. [Natural Earth Dataset](https://www.naturalearthdata.com/)

---


## Code and Reproducibility

The R code used to generate the visualizations in this report is available in 
the repository. Readers can reproduce the analysis by running the scripts provided.

